---
title: "SusannaJenkins_A02_CodingBasics.Rmd"
author: "Susanna Jenkins"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: inline
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on coding basics.

## Directions

1.  Rename this file `<FirstLast>_A02_CodingBasics.Rmd` (replacing `<FirstLast>` with your first and last name).
2.  Change "Student Name" on line 3 (above) with your name.
3.  Work through the steps, **creating code and output** that fulfill each instruction.
4.  Be sure to **answer the questions** in this assignment document.
5.  When you have completed the assignment, **Knit** the text and code into a single PDF file.
6.  After Knitting, submit the completed exercise (PDF file) to Sakai.

## Basics Day 1

1.  Generate a sequence of numbers from one to 100, increasing by fours. Assign this sequence a name.

2.  Compute the mean and median of this sequence.

3.  Ask R to determine whether the mean is greater than the median.

4.  Insert comments in your code to describe what you are doing.

```{r}
#1. seq(1,100,4)
seq <- seq(1,100,4) # this is generating a sequence of numbers from 1 to 100 
#increasing by 4 and then naming it "seq"

#2. 
mean(seq) # this is finding the mean of the sequence
median(seq) # this is finding the median of the sequence
#3. 
mean(seq)>median(seq) # this is showing whether the mean is greater than the 
#median
```

## Basics Day 2

5.  Create a series of vectors, each with four components, consisting of (a) names of students, (b) test scores out of a total 100 points, and (c) whether or not they have passed the test (TRUE or FALSE) with a passing grade of 50.

6.  Label each vector with a comment on what type of vector it is.

7.  Combine each of the vectors into a data frame. Assign the data frame an informative name.

8.  Label the columns of your data frame with informative titles.

```{r}
##5 
#a.
name <- c('susanna', 'irene', 'jordan', 'lucy') # character vector (student 
#names)
testscore <- c(100, 90, 94, 46) # numeric vector (student test scores)
passing <- c(TRUE, TRUE, TRUE, FALSE) # logical vector (TRUE if  score is 
#passing aka greater than 50)

class(testscore)
class(passing)
class(name)

examscores <- data.frame(name=name, testscore=testscore, passing=passing)

examscores
```

9.  QUESTION: How is this data frame different from a matrix?

> Answer: The data frame contains different types of data in the column, but a matrix can only contain a single class of data

10. Create a function with an if/else statement. Your function should take a **vector** of test scores and print (not return) whether a given test score is a passing grade of 50 or above (TRUE or FALSE). You will need to choose either the `if` and `else` statements or the `ifelse` statement. 

11. Apply your function to the vector with test scores that you created in number 5.

```{r}
PassingScores <- function(scores) {
  reviewscores <- ifelse(scores>=50, TRUE, FALSE)
print(reviewscores)
}
PassingScores(testscore)

# PassingScores <- function(scores) {
#   if(scores <= 50) {
#     FALSE
#   } 
#   else {
#     TRUE
#   }
# }
# PassingScores(testscore)
```

12. QUESTION: Which option of `if` and `else` vs. `ifelse` worked? Why?

> Answer: 'ifelse' works here. 'if' 'else' doesn't work here because you can't 
run a vector greater than 1. When I ran 'if' 'else', it gave the following error: 
Error in if (scores <= 50) { : the condition has length > 1
