---
title: "Assignment 5: Data Visualization"
author: "Susanna Jenkins"
date: "Spring 2023"
output: pdf_document
geometry: margin=2.54cm
fig_height: 3
fig_width: 4
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on Data Visualization 

## Directions
1.  Rename this file `<FirstLast>_A05_DataVisualization.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure your code is tidy; use line breaks to ensure your code fits in the knitted output.
5. Be sure to **answer the questions** in this assignment document.
6. When you have completed the assignment, **Knit** the text and code into a single PDF file.

---

## Set up your session 

1. Set up your session. Load the tidyverse, lubridate, here & cowplot packages, and verify your home directory. Upload the NTL-LTER processed data files for nutrients and chemistry/physics for Peter and Paul Lakes (use the tidy `NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv` version) and the processed data file for the Niwot Ridge litter dataset (use the `NEON_NIWO_Litter_mass_trap_Processed.csv` version). 

2. Make sure R is reading dates as date format; if not change the format to date.

```{r initialize project}
#1 
library(tidyverse); library(lubridate); library(here); library(cowplot);

here()

PeterPaul.chem.nutrients <- 
  read.csv(here("Data/Processed_KEY/NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv"), stringsAsFactors = TRUE)

NiwotRidge.litter <- 
  read.csv(here("Data/Processed_KEY/NEON_NIWO_Litter_mass_trap_Processed.csv"), stringsAsFactors = TRUE)

#2 

PeterPaul.chem.nutrients$sampledate <- ymd(PeterPaul.chem.nutrients$sampledate)
NiwotRidge.litter$collectDate <- ymd(NiwotRidge.litter$collectDate)

```
## Define your theme
3. Build a theme and set it as your default theme. Customize the look of at least two of the following: 
 * Plot background
 * Plot title
 * Axis labels
 * Axis ticks/gridlines
 * Legend

```{r build plot theme}
#3

library(ggthemes)

susanna_theme <- theme_base() + 
  theme(
    plot.title = element_text(
      color='purple'
    ),
    plot.background = element_rect(
      color = 'blue',
      fill = 'grey'
    ),
    legend.position = 'right'
    )

```

## Create graphs
For numbers 4-7, create ggplot graphs and adjust aesthetics to follow best practices for data visualization. Ensure your theme, color palettes, axes, and additional aesthetics are edited accordingly.

4. [NTL-LTER] Plot total phosphorus (`tp_ug`) by phosphate (`po4`), with separate aesthetics for Peter and Paul lakes. Add a line of best fit and color it black. Adjust your axes to hide extreme values (hint: change the limits using `xlim()` and/or `ylim()`).

```{r plot total P vs PO4}
#4 

Plot4 <- 
  ggplot(PeterPaul.chem.nutrients, aes(
    x = tp_ug, 
    y = po4,
    color = lakename,
    shape = lakename)) + 
  geom_point() +
  facet_wrap(vars(lakename)) + 
  geom_smooth(method=lm, color = "black") +
  xlim(0, 85) +
  ylim(0, 15) +
  labs(x="Total Phosphorous", y="Phosphate", title = "Total Phosphorus by Phosphate in Peter + Paul Lakes", color = "Lake")
print(Plot4)

```


5. [NTL-LTER] Make three separate boxplots of (a) temperature, (b) TP, and (c) TN, with month as the x axis and lake as a color aesthetic. Then, create a cowplot that combines the three graphs. Make sure that only one legend is present and that graph axes are aligned. 

Tip: R has a build in variable called `month.abb` that returns a list of months;see <https://r-lang.com/month-abb-in-r-with-example>

```{r Create boxplots}
#5 

#(a)

TemperaturePlot <- 
  ggplot(PeterPaul.chem.nutrients, aes(x = factor(month, levels = 1:12, labels=month.abb), y = temperature_C)) +
  geom_boxplot(aes(color = lakename)) +
  labs(
    x = "month",
    y = 'Temperature',
    title = 'Temperature',
    color = 'lakename'
  ) + theme(legend.position = "top")
print(TemperaturePlot)

#(b)

TPPlot <-
  ggplot(PeterPaul.chem.nutrients, aes(x = factor(month, levels = 1:12, labels=month.abb), y = tp_ug)) +
  geom_boxplot(aes(color = lakename)) +
  labs(
    x = "month",
    y = 'Total P',
    title = 'Total P',
    color = 'lakename') + theme(legend.position =  "none")
print(TPPlot)

#(c)

TNPlot <-
  ggplot(PeterPaul.chem.nutrients, aes(x = factor(month, levels = 1:12, labels=month.abb), y = tn_ug)) +
  geom_boxplot(aes(color = lakename)) +
  labs(
    x = "month",
    y = 'Total N',
    title = 'Total N',
    color = 'lakename'
  )  + theme(legend.position =  "none")
print(TNPlot)

library(cowplot)

plot_grid(TemperaturePlot,TPPlot, TNPlot, nrow = 3, align = 'h')

```



Question: What do you observe about the variables of interest over seasons and between lakes? 

> Answer: 

6. [Niwot Ridge] Plot a subset of the litter dataset by displaying only the "Needles" functional group. Plot the dry mass of needle litter by date and separate by NLCD class with a color aesthetic. (no need to adjust the name of each land use)

7. [Niwot Ridge] Now, plot the same plot but with NLCD classes separated into three facets rather than separated by color. 

```{r Plot litter}
#6

PlotLitter <- 
  ggplot(subset(NiwotRidge.litter, functionalGroup == "Needles"), aes(
    x = collectDate, 
    y = dryMass,
    color = nlcdClass)) + 
  geom_point() +
  labs(
    x = "Collection Date",
    y = 'Dry Mass - Needles',
    title = 'Dry Mass of Needle Litter by Date', 
    color='NLCD Class') +
  ylim(0, 20) 
print(PlotLitter)

#7

PlotLitter_Facets <- 
  ggplot(subset(NiwotRidge.litter, functionalGroup == "Needles"), aes(
    x = collectDate, 
    y = dryMass)) + 
  geom_point() +
  labs(
    x = "Collection Date",
    y = 'Dry Mass - Needles',
    title = 'Dry Mass of Needle Litter by Date', 
    color='NLCD Class') +
  ylim(0, 20) +
  facet_wrap(vars(nlcdClass))
print(PlotLitter_Facets)

```
Question: Which of these plots (6 vs. 7) do you think is more effective, and why? 

> Answer: I personally find the first plot (6) more effective because it is 
allows for more visually obvious benchmarking of the litter collection by year 
between the 3 different land cover classes. Color helps me significantly.